---
title: "Process data"
author: "Benjamin Allévius"
date: "`r Sys.Date()`"
output: 
  markdowntemplates::skeleton:
    fig_caption: yes
    fig_width: 7
    fig_height: 5
    toc: true
    toc_depth: 3
bibliography: ../references.bib
---

## Process the brain cancer data
This file shows how to reshape the New Mexico brain cancer data from the
R package _rsatscan_ [@rsatscan] to a format suitable for the _scanstatistics_
package.

We begin by loading the needed packages and the data, which comes as these three
data frames:

* `NM_cas`: contains the number of cancer cases per year, county, age group, and 
  sex. Two counties are missing from this data frame:
    + Cibola: this county was split from Valencia in 1981, so cases in Cibola
      have been counted to Valencia in the data. Cibola is not present in the
      data frames below either.
    + Harding: this county is present in the two other data frames. We will
      replace the case values in Harding with 0, which is the assumed number of
      cases since this county has such a small population. 
* `NM_pop`: with the exception of Cibola county, this data frame contains the 
  population count for all counties, age groups, and gender category in the 
  years 1973, 1982, and 1991.
* `NM_geo`: contains the longitude and latitude of each county's seat 
  (administrative center). Since the coordinates provided by the `rsatscan` 
  package is not in the needed format, we will grab the coordinates from
  [this Wikipedia article](https://en.wikipedia.org/wiki/List_of_counties_in_New_Mexico) 
  instead, and replace this data frame.

```{r load_popcas_data, cache=FALSE}
library(rsatscan)
library(data.table)

NM_pop <- as.data.table(rsatscan::NMpop)
NM_cas <- as.data.table(rsatscan::NMcas)
```

To get more familiar with the counties of New Mexico, let's plot them on a map:

```{r plot_centroids, cache=FALSE}
library(ggplot2)
library(magrittr)
library(dplyr)

# Grab polygon data for plotting
NM_map <- map_data("county", "new mexico")

# Calculate geographical centroids from a polygon
# See: https://en.wikipedia.org/wiki/Centroid#Centroid_of_polygon
polygon_centroid <- function(x0, y0) {
  x1 <- c(x0[-1], x0[1])
  y1 <- c(y0[-1], y0[1])
  A <- sum(x0 * y1 - x1 * y0) / 2
  Cx <- sum((x0 + x1) * (x0 * y1 - x1 * y0)) / (6 * A)
  Cy <- sum((y0 + y1) * (x0 * y1 - x1 * y0)) / (6 * A)
  c(long = Cx, lat = Cy)
}
pgc <- function(df) polygon_centroid(as.numeric(df$long), as.numeric(df$lat))

centroids <- data.frame(subregion = unique(NM_map$subregion), long = NA, lat = NA)

# Calculate geographic centroids for each county
for (i in 1:nrow(centroids)) {
  centroids[i, c("long", "lat")] <- NM_map %>%
    filter(subregion == centroids$subregion[i]) %>%
    pgc
}

# Plot map with labels at centroids
ggplot() + 
  geom_polygon(data = NM_map,
               mapping = aes(x = long, y = lat, group = group),
               color = "white", fill = "grey") +
  geom_text(data = centroids, 
            mapping = aes(x = long, y = lat, label = subregion)) +
  ggtitle("Counties of New Mexico") +
  theme_bw()
```

Before proceeding, we will make some changes to the data to ensure that the 
counties are named the same across all tables. The main functions of the
scanstatistic package require locations to be encoded as integers.
Thus we will make the county names into a factor variable, which are integer 
vectors with additional attributes. All levels of the factor should also be
present in the data.

```{r format_counties, cache=FALSE}
library(stringr)
# library(plyr, warn.conflicts = FALSE)

# This function corrects inconsistencies in the county names
format_counties <- function(x) {
  x %>%
    tolower %>%
    str_replace_all(pattern = " ", replacement = "") %>%
    str_replace_all(pattern = "ñ", replacement = "n") %>%
    str_replace_all(pattern = "guadelupe", replacement = "guadalupe")
}

# Make the county factor variable
all_counties <- NM_map$subregion %>%
  unique %>%
  format_counties %>%
  as.factor

# Character vector without Cibola county
counties <- levels(all_counties)[-which(levels(all_counties) == "cibola")]

# This function corrects inconsistencies in x and makes it into a factor
factorize_counties <- function(x, fac = counties) {
  x %>%
    format_counties %>%
    factor(levels = c(fac, "cibola"))
}

# Fix inconsistencies and factorize counties in the data
centroids %<>% 
  mutate(county = format_counties(subregion),
         county = factorize_counties(county),
         center_long = long, center_lat = lat,
         long = NULL, lat = NULL,
         subregion = NULL)
NM_cas %<>% mutate(county = factorize_counties(county)) %>% as.data.table
NM_pop %<>% mutate(county = factorize_counties(county))
NM_map %<>% mutate(county = factorize_counties(subregion)) %>% as.data.table
NM_pop %<>% mutate(year = year + 1900) %>% as.data.table
```

Scan statistics aim to find _localized_ anomalies in the data, meaning that the 
locations and timepoints involved in the anomaly are close together in some way.
In this example, we will measure the spatial similarity using the spherical 
distance between the county seats. In other circumstances, we might use the
geographic centroids (calculated above), or another distance measure.
So let's grab the coordinates of the county seats from Wikipedia and plot them.
Kudos to Cory Nissen for 
[this blogpost](http://blog.corynissen.com/2015/01/using-rvest-to-scrape-html-table.html), 
which shows how to fetch data from a table on Wikipedia.

```{r grab_seat_coords, cache=FALSE}
library(rvest)

# Function to clean table entries
clean_entry <- function(x) {
  x %>%
    gsub("\\d+♠", "",  .) %>% 
    gsub("^[^\\(]+\\(", "", .) %>%
    gsub("\\,", "", .) %>%
    str_split(pattern = "[:blank:]") %>%
    lapply(function(x) x[1]) %>%
    unlist %>%
    as.numeric
    # gsub("\\ [:alpha:]+", "", .)
    # gsub("\ km2\\)", "", .) %>%
    # gsub("\ km²\\)", "", .)
}

# Grab the table from Wikipedia (as at 2017-08-01)
seat_url <- "https://en.wikipedia.org/wiki/List_of_counties_in_New_Mexico"
NM_geo <- seat_url %>%
  read_html %>%
  rvest::html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[3]') %>%
  html_table %>%
  .[[1]] %>%
  select(County, `County seat[3]`, `Population[6]`, `Area[3][7]`) %>%
  rename(county = County, 
         seat = `County seat[3]`, 
         population = `Population[6]`, 
         `area(km2)` = `Area[3][7]`) %>%
  mutate_at(.vars = c("population", "area(km2)"), .funs = clean_entry) %>%
  mutate(county = gsub(" County", "", county)) %>%
  mutate(county = factorize_counties(county, counties)) %>%
  as.data.table


# Get the coordinates for the New Mexico county seats using ggmap::geocode
get_NM_longlat <- function(seats) {
  suppressMessages({
    res <- t(sapply(seats, function(x) t(ggmap::geocode(paste0(x, ", New Mexico")))))
  })
  data.frame(seat_long = res[, 1], seat_lat = res[, 2])
}

NM_geo <- cbind(NM_geo, get_NM_longlat(NM_geo$seat))
NM_geo <- merge(NM_geo, centroids, by = "county")

ggplot() + 
  geom_polygon(data = NM_map,
               mapping = aes(x = long, y = lat, group = group),
               color = "white", fill = "grey") +
  geom_point(data = NM_geo, 
             mapping = aes(x = seat_long, y = seat_lat)) +
  ggtitle("County seats of New Mexico") +
  theme_bw()
```

For the scan statistic functions in this package to work, locations and 
timepoints with zero counts must be filled in as such, and not left out.
Thus, we will add the needed zeros to the case data. We will only consider the 
total number of cancer cases in each county and year, so we also take the 
oppotunity to aggregate the case counts over each age group and sex:
```{r merge_data, cache=FALSE}
# Aggregate population and cases over sex and age group
NM_cas <- NM_cas[, .(count = sum(cases)), by = .(year, county)]
NM_pop <- NM_pop[, .(population = sum(population)), by = .(year, county)]

# Create the complete, sorted table of counties, years, age groups, and sex
combo_YC <- as.data.table(expand.grid(year = unique(NM_cas$year), 
                                      county = counties))

# Set keys for merge
setkeyv(NM_cas, c("year", "county"))
setkeyv(NM_pop, c("year", "county"))
setkeyv(combo_YC, c("year", "county"))

# Merge data
NM_cas <- merge(NM_cas, combo_YC, all = TRUE)
NM_pop <- merge(NM_pop, combo_YC, all = TRUE)

# Fill in zero case counts
NM_cas[is.na(count), count := 0]
```

To demonstrate the scan statistic methods of this package, we will try to detect
brain cancer clusters in the years 1986--1989, using data from the previous 
years to predict how many cancer cases we can expect to see if there is no 
anomalous increase in incidence. In this particular example, we want to use the
available population size data as offsets to our predicted means, in order to
properly compare the occurence of cancer in the different counties. Since the 
population counts are only available for three of the years, we will interpolate 
the population for the remaining years by fitting a quadratic function of time 
to the population data for each county after aggregating the number of cancer
cases over age group and gender[^1]. Though this leads to unrealistically smooth
population changes, it is adequate for the purpose of demonstrating the 
functions in the _scanstatistics_ package.

[^1]: One could also fit the quadratic function for each county, age group and
sex, and then aggregate over the latter two categories in the hope that this 
would cancel out the errors made in each interpolation. However, this seems to
lead to some unreasonable year-to-year population changes for some of the 
counties.
```{r predict_pop, cache=FALSE}
NM_pop[, population := as.numeric(population)]

# Interpolate county populations
for (s in counties) {
  pop_mod <- lm(population ~ 1 + I(scale(year)) + I(scale(year)^2), 
                data = NM_pop[county == s, ])
  pop_pred <- predict(pop_mod, NM_pop[is.na(population) & county == s, ], 
                      type = "response")
  NM_pop[is.na(population) & county == s, population := pop_pred]
}

```

Finally, we join the population and case tables and plot how the number of
cases evolve in each county over the years:
```{r plot_evolution, cache=FALSE}
NM_popcas <- merge(NM_pop, NM_cas)
setkeyv(NM_popcas, c("county", "year"))

ggplot(NM_popcas) + geom_line(aes(x = year, y = count, color = county)) +
  theme(legend.position = "none")
```

The data tables `NM_popcas`, `NM_geo` and `NM_map` are then saved as those 
available in the _scanstatistics_ package, after converting them to data frames:

```{r convert_tables}
NM_popcas %<>% as.data.frame
NM_geo %<>% as.data.frame
NM_map %<>% as.data.frame
```
